Spatial variation of the native colon microbiota in healthy adults

### 2017-11-14 submission/manuscript.Rmd
version sent to collabs
first submission draft complete

---
title: "Spatial variation of the native colon microbiota in healthy adults"
csl: aacr.csl
fontsize: 12pt
output:
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
  word_document: default
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \renewcommand{\familydefault}{\sfdefault}
- \usepackage{graphicx}
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{8mm}

@@

Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, D. Kim Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^

@@

The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location. Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017]. The manifestations of these diseases are known to vary based upon the location in which they occur. For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012. In contrast, CRC arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012. In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015]. In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016. UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbiota and their metabolites likely vary as well, and may influence the heterogeneous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbiota is important to understanding the pathophysiology of both diseases.

@@

### Methods
#### Human subjects
The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721. Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study. Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl or Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive. 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure. There were 13 female and 7 male subjects ranging in age from 25 to 64.

@@

#### Data availability
16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number XXXX. A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017.

@@

### Acknowledgments 
We thank all the individuals who volunteered for the study. This work was supported by the Rose and Lawrence C. Page Foundation (DKT). We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing.

>>>>

---
title: "Spatial variation of the native colon microbiota in healthy adults"
csl: aacr.csl
fontsize: 12pt
output:
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
  word_document: default
sansfont: Arial
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
- \renewcommand{\familydefault}{\sfdefault}
- \usepackage{graphicx}
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{8mm}

@@

Kaitlin J. Flynn^1^, Mack T. Ruffin IV^2^, D. Kim Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^

@@

The human colon is an ecosystem comprised of numerous microenvironments that select for different microbiota. Concentrations of oxygen, water, and anti-microbial peptides change along the gut axis and influence which microbiota reside in each location. Microenvironments differ not only longitudinally along the colon, but also radially from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbiota and communities are important for understanding the etiology of complex diseases such as Colorectal Cancer (CRC) and Inflammatory Bowel Disease (IBD). CRC and IBD can be preceded or accelerated by perturbations the structure of the gut microbiota [@Yamauchi2012; @Forbes2016; @Halfvarson2017]. The manifestations of these diseases are known to vary based upon the location in which they occur. For instance, CRC that arises in the distal (left) colon are of hindgut origin and tend to have large chromosomal alterations indicative of chromosomal instability @Yamauchi2012. In contrast, CRC arising in the proximal (right) colon tumors are of midgut origin and tend to be sessile and microsatellite instable (MSI with BRAF and KRAS mutations) @Yamauchi2012. In addition to the environmental gradients within the colon, the distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively [@Yamauchi2012; @Albenberg2014; @Donaldson2015]. In IBD patients, disease occurring in the distal colon extending proximally is usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) can occur anywhere along the GI tract, most commonly in the ileum and the cecum @Forbes2016. UC presents as continuous disease with only mucosal involvement, where as CD has skip lesions and full thickness involvement that may cause abscesses, strictures and fistulas @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbiota and their metabolites likely vary as well, and may influence the heterogeneous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbiota is important to understanding the pathophysiology of both diseases.

@@

### Methods
#### Human subjects
The procedures in this study and consent were approved by the Institutional Review Board at the University of Michigan Health System with protocol number HUM00082721. Subjects were recruited using the online recruitment platform and were pre-screened prior to enrollment in the study. Exclusion criteria included: use of asprin or NSAIDs within 7 days, use of antibiotics within 3 months, current use of anticoagulants, known allergies to Fentanyl, Versed and Benadryl, prior history of colon disease, diabetes, abdominal surgery, respiratory, liver, kidney or brain impairments, undergoing current chemotherapy or radiation treatment and subjects that were pregnant or trying to conceive. 20 subjects that met the criteria were selected and provided signed informed consent prior to the procedure. There were 13 female and 7 male subjects ranging in age from 25 to 64.

@@

#### Data availability
16S rRNA gene sequence reads and experiment metadata are available on the NCBI Sequence Read Archive (SRA) with accession number SRP124947 and PRJNA418115. A reproducible data analysis pipeline can be found at https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017.

@@

### Acknowledgments 
We thank all the individuals who volunteered for the study. This work was supported by the Rose and Lawrence C. Page Foundation (DKT). We would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman for their roles in study coordination, subject recruitment, sample collection and sample processing. Thanks also to GI Fellows Drs Amy Hosmer, Alexander Larson and Anna Tavakkoli who assisted Dr. Turgeon with the colonoscopies.

### 2017-11-14 submission/manuscript.tex
first submission draft complete

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Spatial variation of the native colon microbiota in healthy adults},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

Kaitlin J. Flynn\textsuperscript{1}, Charles C.
Koumpouras\textsuperscript{1}, Mack T. Ruffin IV\textsuperscript{2}, D.
Kim Turgeon\textsuperscript{3\(\dagger\)}, and Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

@@

The procedures in this study and consent were approved by the
Institutional Review Board at the University of Michigan Health System
with protocol number HUM00082721. Subjects were recruited using the
online recruitment platform and were pre-screened prior to enrollment in
the study. Exclusion criteria included: use of asprin or NSAIDs within 7
days, use of antibiotics within 3 months, current use of anticoagulants,
known allergies to Fentanyl or Benadryl, prior history of colon disease,
diabetes, abdominal surgery, respiratory, liver, kidney or brain
impairments, undergoing current chemotherapy or radiation treatment and
subjects that were pregnant or trying to conceive. 20 subjects that met
the criteria were selected and provided signed informed consent prior to
the procedure. There were 13 female and 7 male subjects ranging in age
from 25 to 64.

@@

16S rRNA gene sequence reads and experiment metadata are available on
the NCBI Sequence Read Archive (SRA) with accession number XXXX. A
reproducible data analysis pipeline can be found at
\url{https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017}.

@@

We thank all the individuals who volunteered for the study. This work
was supported by the Rose and Lawrence C. Page Foundation (DKT). We
would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
for their roles in study coordination, subject recruitment, sample
collection and sample processing.

>>>>

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setsansfont[]{Arial}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Spatial variation of the native colon microbiota in healthy adults},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

Kaitlin J. Flynn\textsuperscript{1}, Mack T. Ruffin
IV\textsuperscript{2}, D. Kim Turgeon\textsuperscript{3\(\dagger\)}, and
Patrick D. Schloss\textsuperscript{1\(\dagger\)}

@@

The procedures in this study and consent were approved by the
Institutional Review Board at the University of Michigan Health System
with protocol number HUM00082721. Subjects were recruited using the
online recruitment platform and were pre-screened prior to enrollment in
the study. Exclusion criteria included: use of asprin or NSAIDs within 7
days, use of antibiotics within 3 months, current use of anticoagulants,
known allergies to Fentanyl, Versed and Benadryl, prior history of colon
disease, diabetes, abdominal surgery, respiratory, liver, kidney or
brain impairments, undergoing current chemotherapy or radiation
treatment and subjects that were pregnant or trying to conceive. 20
subjects that met the criteria were selected and provided signed
informed consent prior to the procedure. There were 13 female and 7 male
subjects ranging in age from 25 to 64.

@@

16S rRNA gene sequence reads and experiment metadata are available on
the NCBI Sequence Read Archive (SRA) with accession number SRP124947 and
PRJNA418115. A reproducible data analysis pipeline can be found at
\url{https://github.com/SchlossLab/Flynn_LRColon_CancPrevRes_2017}.

@@

We thank all the individuals who volunteered for the study. This work
was supported by the Rose and Lawrence C. Page Foundation (DKT). We
would also like to thank Brian Kleiner, Chelsea Crofoot, and Kirk Herman
for their roles in study coordination, subject recruitment, sample
collection and sample processing. Thanks also to GI Fellows Drs Amy
Hosmer, Alexander Larson and Anna Tavakkoli who assisted Dr.~Turgeon
with the colonoscopies.
